---
title: "Proposal Presentation Outline"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
Research Question: What shared variants/loci exist between severe early-onset allergic phenotypes and T1D, and are these loci underrepresented in Black populations in ClinVar?

The combination of all four elements together — early-onset severe allergic phenotypes, T1D, Black ancestry representation, and LLM-assisted bioinformatics — is something I'm not aware of existing in the literature in any comprehensive form. Each piece individually has some work done on it, but no one has brought them together into a unified study. That's the definition of a research gap worth filling.

My Expectations:
I expect to find a correlated relationship between early-onset severe allergic responses and the subsequent development of Type 1 Diabetes. 

EDA Checklist:
1. Formulate Question
2. Read the data
3. Check the packaging
4. Look at head and tail
5. Check the n's
6. Validate with 1 external source


Load and Read the `variant_summary.txt` into R environment
```{r}
library(tidyverse)
getwd()
setwd("C:/Data Wrangling/Data Wrangling/Final Project")

```
Read Data in tsv
```{r}
Variant_data <- read_tsv("variant_summary.txt", col_names = TRUE)
glimpse(Variant_data)
```
Check the packaging
1. Check for null values 
```{r}
str(Variant_data)
```
```{r}
colSums(is.na(Variant_data))
```
```{r}
sum(is.na(Variant_data))
```
2. Check number of rows 
```{r}
nrow(Variant_data)

```
3. Check number of columns 
```{r}
ncol(Variant_data)
```

Print columns
```{r}
colnames(Variant_data)
```
With the help of Gemini I have identified that my columns a combination of 4 different categories:
1. The Core identifiers and basic info
  - #AlleleID: primary key for a specific allele(specific variant) in CLinVar
  - Type: the type of variant (single nucleotide variant = SNV), deletion, insertion, duplication, etc. 
  - Name: full (human readable name) - often written in HGVS notation
  - GeneID: tells us where the gene is located (NCBI Entrez Gene ID)
  - GeneSymbol: The short official symbol for the gene 
  - HGNC_ID: ID assigned by the HUGO Gene Nomenclature Committee (office authority for human gene naming)
2. Clinical Significance 
  - ClinicalSignificance: full 'clinical' interpretation of the variant (Pathogenic, Benign, Uncertain significance). *if submitters disagree multiple terms with be included
  - ClinSigSimple: Numeric version of ClincalSignificance 
  - LastEvaluated: most recent date that submitter review/evaluated the ClinicalSignificance of that specific variance
  - TeviewStatus: reflects how well supported the classification is ('no assertion criteria provided' to 'practice guideline'. More stars = More confidence)
  - NumberSubmitters: reflects how many independent labs or organizations have submitted evidence about this variant to ClinVar
  - Guidelines: list any 'official' clinicalguidelines that reference this variant
3. Disease/Phenotype
  - PhenotypeIDS: Database IDS (from MedGen, OMIM, Orphanet) for the diseases(s) or conditions(s) associated with the variant
  - PhenotypeList: The names of the conditions
  - Origin: where the variant came from (germiline=inherited, somatic=acquired in a tumor, de novo=new mutation)
  - OriginSimple: simplified version of the origin
4. Genomic Location
  - Assembly: the human genome reference assembly 
  - ChromosomeAccession: the official NCBIS acession number match to the specific chromosome sequence used
  - Chromosome: the chromosome where the variant is located 
  - Start: the starting genomic coordinate of the variant on the chromosome
  - Stop: the ending genomic coordinate (same as the start for a single nucleotide variants)
  - Cytogenetic: the cytogenetic band location (used in karyotyping)
  - ReferenceAllele: the nucleotide(s) found in the reference genome at the position
  - AlternateAllele: the nucleotide(s) that differ from the references (the actual variant sequence)
5. Cross-Database References
  - RS#(dbSNP): the number from the NCBI's database of known variants
  - nsv/esv(dbVAR): structural variant ID from dbVar, NCBI's database for larger genomic variants (i.e. copy number variants)
  - RCVaccession: tied to a specific variant-condition pair (one variant can have many RVC's for different diseases) 
  - VariationID: this ID is independent of the condition. Used ot look it up directly on the CinVar website
  - OtherIDS: Additional cross-references to other databases
6. VCF Fields
  - PositionVCF:chromosomal position as it would appear in a standard VCF (Variant Call Format) file
  - ReferenceAlleleVCF: reference allele formatted specifically for VCF standards 
  - AlternateAlleleVCF: alternate allele formatted for VCF standards
7. Submission & Testing
  - TestedInGTR: Y or N for whether this variant is included in a genetic test registered in NCBI's Genetic Testing Registry (GTR)
  - SubmitterCategories: the types of organizations that submitted data (clinical laboratory, research, expert panel)
8. Somatic/Oncogenicity Fields
  - SomaticClinicalImpact: the clinical impact of the variant when found in a tumor (somatically) - like "Tier 1 - Strong Clinical Significance"
  - SomaticClinicalImpactLastEvaluated: the date the somatic clinical impact was last reviewed
  - ReviewStatusClinicalImpact: the confidence level specifically for the somatic clinical impact classification
  - Oncogenicity: is the variant consider 'cancer-causing' (oncogenic) when its a tumor (somatic)
  - OncogenicityLastEvaluated: the date the oncogenicity classification was last reviewed
  - ReviewStatusOncogenicity: confidence level for the oncogenicity classification
9. AggregateSCVFields
  - SCVsForAggregateGermlineClassification: the list of individual SCVs that contributed to the overall germline pathogenicity classification
  - SVCsForAgreggateSomaticClinicalImpact: the list of SCVs contributing to the somatic clinical impact classification
  - SCVSForAggregatedOncogenicityClassification: the list of SCVs contributing to the oncogenicity classification




Print the head
```{r}
head(Variant_data)
```

Print the tail
```{r}
table(sapply(Variant_data, class))
```
Filter down to type 1 data using Keywords: type 1 diabetes, T1D, IDDM, insulin-dependent diabetes

Widen the search and then also filtered to find genesymbols associated with type 1 diabetes to further sift out any noise within the data
```{r}
t1d_variants <- Variant_data %>%
  filter(str_detect(PhenotypeList, regex("type 1 diabetes|T1D|IDDM|insulin-dependent diabetes|diabetes mellitus|type 1 diabetes mellitus 20|monogenic autoimmune diabetes|autoimmune diabetes|neonatal diabetes mellitus|immune related diabetes|MODY|maturity-onset diabetes of the young|maturity onset diabetes mellitus in young", ignore_case = TRUE))) %>% filter(str_detect(GeneSymbol, regex("HLA-DRB1|HLA-DQA1|HLA-DQB1|INS|CTLA4|PTPN22|HNF1A|GCK|HNF4A", ignore_case = TRUE)))

# Check how many variants 
nrow(t1d_variants)

# Preview
glimpse(t1d_variants)
```

Check the Phenotypes and their counts <----- Human Version Results
```{r}
t1d_variants %>%
  count(PhenotypeList, sort = TRUE)
```

Severe Anaphalaxus filter
```{r}
anaphylaxis <- Variant_data %>%
  filter(str_detect(PhenotypeList, regex("mastocytosis|Hereditary alpha-tryptasemia|HαT|Idiopathic anaphylaxis|Mast cell activation syndrome", ignore_case = TRUE))) %>% filter(str_detect(GeneSymbol, regex("KIT|TPSAB1|FCER1A|(IL4&IL3)", ignore_case = TRUE)))

# Check how many variants 
nrow(anaphylaxis)

# Preview
glimpse(anaphylaxis)
```
```{r}
anaphylaxis %>%
  count(PhenotypeList, sort = TRUE)
```
Official Trait List:
- Phenotypes:
    Anaphylaxis: 
      mastocytosis
      Hereditary alpha-tryptasemia
      HαT
      Idiopathic anaphylaxis|Mast cell activation syndrome
    T1D: 
      type 1 diabetes
      T1D
      IDDM
      insulin-dependent diabetes
      diabetes mellitus
      type 1 diabetes mellitus 20
      monogenic autoimmune diabetes
      autoimmune diabetes|neonatal diabetes mellitus
      immune related diabetes|MODY|maturity-onset diabetes of the young|maturity onset diabetes mellitus in young
- GeneSymbols:
    Anaphylaxis: 
      KIT
      TPSAB1
      FCER1A
      IL4&IL3
    T1D: 
      HLA-DRB1
      HLA-DQA1
      HLA-DQB1
      INS
      CTLA4
      PTPN22
      HNF1A
      GCK
      HNF4A

Repeat filter process by via LLM query 
```{r}

```


Download GWAS acession dataset for ...
```{r}
getwd()
gwas <- read_tsv("C:/Data Wrangling/Data Wrangling/Final Project/gwas-catalog-download-associations-alt-full.tsv",
                 col_types = cols(.default = "c"))
glimpse(gwas)
table(sapply(gwas, class))
sum(is.na(gwas))

colnames(gwas)
```

Preform tidy data task
```{r}
head(gwas)
```
Download ancestor GWAS dataset to get the demogrpahic information of the studies. 

```{r}
gwas_ancestor <- read_tsv("C:/Data Wrangling/Data Wrangling/Final Project/gwas-catalog-v1.0.3.1-ancestries-r2026-02-16.tsv",
                 col_types = cols(.default = "c"))
glimpse(gwas_ancestor)
gwas_ancestor %>%
  group_by(`BROAD ANCESTRAL CATEGORY`)%>%
  count()
(gwas_ancestor)
table(sapply(gwas, class))

sum(is.na(gwas))
```
Filter for GWAS studies that atleast inclue "African American or Afro-Caribbean|African unspecified" in the Broad Ancestral Category 

```{r}
library(stringr)

ancestry_processed <- gwas_ancestor %>%
  mutate(has_african_ancestry = str_detect(
    `BROAD ANCESTRAL CATEGORY`, 
    "African American or Afro-Caribbean|African unspecified"
  ))

# filter your main dataset for these studies
african_ancestry_studies <- ancestry_processed %>%
  filter(has_african_ancestry == TRUE)

glimpse(african_ancestry_studies)
```
Filter for observations that ONLY include black participants.

```{r}
gwas_ancestor <- gwas_ancestor%>% filter(`BROAD ANCESTRAL CATEGORY`== 'African American or Afro-Caribbean' | `BROAD ANCESTRAL CATEGORY`== 'African American' | `BROAD ANCESTRAL CATEGORY`== 'African unspecified'|`BROAD ANCESTRAL CATEGORY`== 'Afro-Caribbean' |`BROAD ANCESTRAL CATEGORY`== 'Sub-Saharan African'| `BROAD ANCESTRAL CATEGORY`== 'African unspecified, African American or Afro-Caribbean'| `BROAD ANCESTRAL CATEGORY`=='African American or Afro-Caribbean, NR'| `BROAD ANCESTRAL CATEGORY`== 'African American or Afro-Caribbean, African unspecified'|`BROAD ANCESTRAL CATEGORY`=='	African unspecified, NR'|`BROAD ANCESTRAL CATEGORY`=='NR, African unspecified'|`BROAD ANCESTRAL CATEGORY`== '	Sub-Saharan African, African American or Afro-Caribbean'|`BROAD ANCESTRAL CATEGORY`== 'Sub-Saharan African, African unspecified')

glimpse(gwas_ancestor)
```
Join both GWAS datasets 
```{r}
joined_gwas <- gwas_ancestor %>%
  inner_join(gwas)
joined_gwas

colnames(joined_gwas)

glimpse(joined_gwas)

table(sapply(joined_gwas, class))

sum(is.na(gwas))
```

```{r}
#t1d filter
t1d_subset <- joined_gwas %>% filter(str_detect(`DISEASE/TRAIT`, "type 1 diabetes"))
#Severe allergy filter
allergy_subset <- joined_gwas %>% filter(str_detect(`DISEASE/TRAIT`, "mastocytosis|anaphylaxis|allergy"))

# Find shared variants
shared_loci <- inner_join(t1d_subset, allergy_subset, by = "SNPS")

glimpse(t1d_subset)
glimpse(allergy_subset)
shared_loci
```
So here I found that in my current filter statements there is no overlap. I am going to broaden my filters 

SHOW THIS TO DR.AZIZ. ASK IF ITS NECESSARY TO INCLUDE THE VARCHAR DATA???

```{r}
# Expanded the search
#t1d filter
t1d_subset <- joined_gwas %>% filter(str_detect(`DISEASE/TRAIT`, "type 1 diabetes|T1D|IDDM|insulin-dependent diabetes|diabetes mellitus|type 1 diabetes mellitus 20|monogenic autoimmune diabetes|autoimmune diabetes|neonatal diabetes mellitus|immune related diabetes|MODY|maturity-onset diabetes of the young|maturity onset diabetes mellitus in young"))
#severe allergy filter
allergy_subset <- joined_gwas %>% filter(str_detect(`DISEASE/TRAIT`, "mastocytosis|Hereditary alpha-tryptasemia|HαT|Idiopathic anaphylaxis|Mast cell activation syndrome|mastocytosis|anaphylaxis|allergy"))

# Find shared variants
shared_loci <- inner_join(t1d_subset, allergy_subset, by = "SNPS")

glimpse(t1d_subset)
glimpse(allergy_subset)
shared_loci
```
I am plan to either us Gemini Flash or Ollama for my LLM backend since both are free.

##One Time Intergration Step
```{r}
library(httr2)
library(jsonlite)

# Using Google Gemini free tier
llm_call <- function(prompt, api_key = Sys.getenv("GEMINI_API_KEY")) {
  req <- request("https://generativelanguage.googleapis.com/v1beta/models/gemini-1.5-flash:generateContent") |>
    req_url_query(key = api_key) |>
    req_headers("Content-Type" = "application/json") |>
    req_body_json(list(
      contents = list(list(parts = list(list(text = prompt))))
    )) |>
    req_perform()
  
  resp_body_json(req)$candidates[[1]]$content$parts[[1]]$text
}
```

##Feature Type Inference on variant_summary columns
```{r}
infer_variant_column_types <- function(variant_df) {
  # Sample a few values per column to keep prompt concise
  col_samples <- sapply(names(variant_df), function(col) {
    vals <- na.omit(variant_df[[col]])[1:3]
    paste0(col, ": [", paste(vals, collapse = ", "), "]")
  })
  
  prompt <- paste0(
    "You are analyzing a ClinVar variant_summary genomic dataset.\n",
    "For each column below, classify it as one of: Numerical, Categorical, ",
    "Datetime, Ignorable_ID, Formatted_ID, Sentence, URL, List, or Text.\n",
    "Also flag columns likely 'influential' for T1D or allergy phenotype analysis.\n",
    "Return only JSON: [{\"column\": \"...\", \"type\": \"...\", \"influential\": true/false}]\n\n",
    paste(col_samples, collapse = "\n")
  )
  
  result <- llm_call(prompt)
  fromJSON(gsub("```json|```", "", result))
}

# Usage
fti_results <- infer_variant_column_types(variant_summary_df)
influential_cols <- fti_results$column[fti_results$influential == TRUE]
```

##Phentoype Labeling - Semantic matching for T1D & Food related Anaphilaxis
```{r}
semantic_phenotype_match <- function(clinvar_terms, gwas_terms) {
  prompt <- paste0(
    "You are a biomedical ontology expert.\n",
    "ClinVar phenotype terms: ", paste(clinvar_terms, collapse = "; "), "\n",
    "GWAS phenotype terms: ", paste(gwas_terms, collapse = "; "), "\n\n",
    "Identify semantic matches relevant to Type 1 Diabetes OR severe allergic ",
    "reactions (anaphylaxis, food allergy, allergic rhinitis, eczema, asthma).\n",
    "Include near-matches that rigid string filters would miss.\n",
    "Return only JSON: [{\"clinvar_term\": \"...\", \"gwas_term\": \"...\", ",
    "\"relevance\": \"T1D|Allergy|Both\", \"confidence\": 0.0-1.0}]\n"
  )
  
  result <- llm_call(prompt)
  fromJSON(gsub("```json|```", "", result))
}

# unique phenotype values
matched_phenotypes <- semantic_phenotype_match(
  clinvar_terms = unique(clinvar_subset$PhenotypeList),
  gwas_terms    = unique(gwas_subset$DISEASE.TRAIT)
)

# Filter high-confidence matches
strong_matches <- matched_phenotypes[matched_phenotypes$confidence >= 0.75, ]
```
##Demograpic Filtering - Quantifying Black participants in GWAS
```{r}
classify_gwas_ancestry <- function(gwas_df) {
  # GWAS ancestry descriptions are often inconsistent strings
  unique_ancestry <- unique(gwas_df$BROAD.ANCESTRAL.CATEGORY)
  
  prompt <- paste0(
    "You are analyzing GWAS study ancestry classifications.\n",
    "For each ancestry label below, determine:\n",
    "1. Whether it includes Black/African/African-American participants\n",
    "2. A standardized category: African, European, Asian, Mixed, Unknown, Other\n",
    "Labels: ", paste(unique_ancestry, collapse = "; "), "\n\n",
    "Return only JSON: [{\"original\": \"...\", \"includes_black\": true/false, ",
    "\"standardized\": \"...\"}]\n"
  )
  
  result <- llm_call(prompt)
  mapping <- fromJSON(gsub("```json|```", "", result))
  
  # Join back to full dataset
  gwas_df |>
    dplyr::left_join(mapping, by = c("BROAD.ANCESTRAL.CATEGORY" = "original"))
}

# Usage
gwas_classified <- classify_gwas_ancestry(gwas_subset)

# Summarize Black representation
black_studies <- gwas_classified |>
  dplyr::filter(includes_black == TRUE) |>
  dplyr::summarise(
    n_studies       = dplyr::n_distinct(STUDY.ACCESSION),
    n_associations  = dplyr::n()
  )
```

##Data Joining - GWAS + ClinVar via RSID and Chromosomal Position
```{r}
llm_assisted_join <- function(unmatched_gwas, unmatched_clinvar) {
  # For rows where RSID join failed, try LLM-guided positional reconciliation
  sample_gwas    <- head(unmatched_gwas[, c("SNP", "CHR_ID", "CHR_POS")], 20)
  sample_clinvar <- head(unmatched_clinvar[, c("RS.", "Chromosome", "Start")], 20)
  
  prompt <- paste0(
    "These genomic variants failed to join by RSID. ",
    "Suggest join logic using chromosome + base pair position.\n",
    "GWAS columns: SNP (rsid), CHR_ID (chromosome), CHR_POS (position)\n",
    "ClinVar columns: RS# (rsid), Chromosome, Start (position)\n\n",
    "GWAS sample:\n", toJSON(sample_gwas, auto_unbox = TRUE), "\n\n",
    "ClinVar sample:\n", toJSON(sample_clinvar, auto_unbox = TRUE), "\n\n",
    "Return R dplyr join code to reconcile these, accounting for ",
    "position tolerances (±1bp) and chromosome formatting differences (e.g. '1' vs 'chr1').\n"
  )
  
  llm_call(prompt)  # Returns suggested R code you can review and run
}

# Two-stage join strategy
# Stage 1: exact RSID match
joined_exact <- dplyr::inner_join(gwas_subset, clinvar_subset,
                                   by = c("SNP" = "RS."))

# Stage 2: LLM-guided positional join for unmatched rows
unmatched_gwas    <- dplyr::anti_join(gwas_subset, joined_exact, by = "SNP")
unmatched_clinvar <- dplyr::anti_join(clinvar_subset, joined_exact, by = c("RS." = "SNP"))

join_suggestion <- llm_assisted_join(unmatched_gwas, unmatched_clinvar)
cat(join_suggestion)  # Review before executing
```
