##Install and load Gemini into workspace

```{r}
library(httr2)
library(jsonlite)

#install.packages("GenAI")
library(GenAI)
usethis::edit_r_environ()
Sys.getenv("GOOGLE_API_KEY")
# Setup
google <- genai.google(api = "", model ="gemini-2.5-flash", version="v1")

# 3. Use the txt() function instead of gen()
response <- txt(genai.object = google, 
                prompt = "Write a haiku about R programming.")

# 4. View the result
cat(response)


```
##Load Data into workspace

```{r}
library(tidyverse)
Variant_data <- read_tsv("variant_summary.txt", col_names = TRUE)
gwas_accession <- read_tsv("gwas-catalog-download-associations-alt-full.tsv",
                 col_types = cols(.default = "c"))
gwas_ancestor <- read_tsv("gwas-catalog-v1.0.3.1-ancestries-r2026-02-16.tsv",
                 col_types = cols(.default = "c"))
colnames(gwas_accession)
```
## Perform Feature Type Inference on columns in Each dataset

*Vairant_data -> ClinVar
```{r}
library(jsonlite)

# Define the function
infer_variant_column_types <- function(Variant_data, genai_obj) {
  # 1. Sample the data
  col_samples <- sapply(names(Variant_data), function(col) {
    vals <- na.omit(Variant_data[[col]])[1:3]
    if(length(vals) == 0) vals <- "All NA"
    paste0(col, ": [", paste(vals, collapse = ", "), "]")
  })
  
  # 2. Build the prompt (using single quotes for the string to avoid escape errors)
  prompt <- paste0(
    "You are a genomic data expert analyzing a ClinVar dataset.\n",
    "Classify each column as: Numerical, Categorical, Datetime, Ignorable_ID, Formatted_ID, Sentence, URL, List, or Text.\n",
    "CRITICAL: Flag 'influential' as true if the column contains data relevant to identifying risks for:\n",
    "1. Type 1 Diabetes (e.g., HLA alleles, INS, PTPN22 genes).\n",
    "2. Food-related Anaphylaxis/Allergies (e.g., IgE response genes, FLG, or phenotype descriptions).\n",
    "Return only raw JSON: [{\"column\": \"...\", \"type\": \"...\", \"influential\": true/false}]\n\n",
    paste(col_samples, collapse = "\n")
  )
  
  # 3. Call Gemini
  result <- txt(genai.object = genai_obj, prompt = prompt)
  
  # 4. Clean and Parse
  clean_json <- gsub("```json|```", "", result)
  
  tryCatch({
    fromJSON(clean_json)
  }, error = function(e) {
    message("JSON Parsing Failed. Raw output: ", result)
    return(NULL)
  })
}

# --- EXECUTION ---

# 1. Initialize (Paste your API key between the quotes)
google_obj <- genai.google(
  api = api, 
  model = "gemini-2.5-flash", 
  version = "v1"
)

# 2. Run the inference
# Make sure your 'Variant_data' or GWAS data is loaded before running this
fti_results_variant <- infer_variant_column_types(Variant_data, google_obj)

# 3. Filter and Print
if (!is.null(fti_results_variant)) {
  influential_cols <- fti_results_variant$column[fti_results_variant$influential == TRUE]
  cat("Influential Columns Found:\n")
  print(influential_cols)
}
```
 *gwas_accession -> GWAS, Accesssion
 
```{r}
infer_gwas_column_types <- function(GWAS_data, genai_obj) {
  # Sample data for context
  col_samples <- sapply(names(GWAS_data), function(col) {
    vals <- na.omit(GWAS_data[[col]])[1:3]
    if(length(vals) == 0) vals <- "All NA"
    paste0(col, ": [", paste(vals, collapse = ", "), "]")
  })
  
  # Prompt adjusted for GWAS logic
  prompt <- paste0(
    "You are a genomic expert analyzing GWAS summary statistics (accession data).\n",
    "Classify each column as: Numerical, Categorical, Datetime, Ignorable_ID, Formatted_ID, Sentence, URL, List, or Text.\n",
    "Flag 'influential' as true if the column is critical for statistical association analysis for:\n",
    "1. Type 1 Diabetes (T1D)\n",
    "2. Food-related Anaphylaxis/Allergies\n",
    "Prioritize columns containing p-values, effect sizes (OR/beta), rsIDs, or population/ancestry data.\n",
    "Return only raw JSON: [{\"column\": \"...\", \"type\": \"...\", \"influential\": true/false}]\n\n",
    paste(col_samples, collapse = "\n")
  )
  
  # Execute and Clean
  result <- txt(genai.object = genai_obj, prompt = prompt)
  clean_json <- gsub("```json|```", "", result)
  
  tryCatch({
    fromJSON(clean_json)
  }, error = function(e) {
    message("JSON Error. Raw output: ", result)
    return(NULL)
  })
}

# 1. Initialize your object (ensure you use your new key)
google_obj <- genai.google(
  api = api, 
  model = "gemini-2.5-flash", 
  version = "v1"
)

# 2. Run the inference on your variant data
# Assuming Variant_data is already loaded in your environment
fti_results_accession <- infer_gwas_column_types(gwas_accession, google_obj)

# 3. Filter for the influential columns
if (!is.null(fti_results_accession)) {
  influential_cols <- fti_results_accession$column[fti_results_accession$influential == TRUE]
  print(influential_cols)
}
```
 *gwas_ancestory -> GWAS, Ancestry
 
```{r}
infer_ancestor_column_types <- function(gwas_ancestor, genai_obj) {
  # Sample a few values per column to keep prompt concise
  col_samples <- sapply(names(gwas_ancestor), function(col) {
    vals <- na.omit(gwas_ancestor[[col]])[1:3]
    # Handle cases where there might be fewer than 3 non-NA values
    if(length(vals) == 0) vals <- "All NA"
    paste0(col, ": [", paste(vals, collapse = ", "), "]")
  })
  
  prompt <- paste0(
    "You are a genomic expert analyzing GWAS (Genome-Wide Association Study) summary statistics.\n",
    "Classify each column as: Numerical, Categorical, Datetime, Ignorable_ID, Formatted_ID, Sentence, URL, List, or Text.\n",
    "CRITICAL: Flag 'influential' as true if the column is essential for identifying genetic associations for:\n",
    "1. Type 1 Diabetes (T1D)\n",
    "2. Food-related Anaphylaxis/Allergies\n",
    "3. Population-specific data (e.g., Ethnicity, Ancestry, African/Black population markers).\n",
    "Prioritize columns containing: P-values, Beta coefficients, Odds Ratios (OR), rsIDs, and Sample Ancestry.\n",
    "Return only raw JSON: [{\"column\": \"...\", \"type\": \"...\", \"influential\": true/false}]\n\n",
    paste(col_samples, collapse = "\n")
  )
  
  # Use the working GenAI txt() function
  result <- txt(genai.object = genai_obj, prompt = prompt)
  
  # Clean the response string (removes markdown code blocks if the LLM adds them)
  clean_json <- gsub("```json|```", "", result)
  
  # Convert to data frame
  tryCatch({
    fromJSON(clean_json)
  }, error = function(e) {
    message("Error parsing JSON from LLM. Raw output was: ", result)
    return(NULL)
  })
}

# 1. Initialize your object (ensure you use your new key)
google_obj <- genai.google(
  api = api, 
  model = "gemini-2.5-flash", 
  version = "v1"
)

# 2. Run the inference on your variant data
# Assuming Variant_data is already loaded in your environment
fti_results_ancestor <- infer_ancestor_column_types(gwas_ancestor, google_obj)

# 3. Filter for the influential columns
if (!is.null(fti_results)) {
  influential_cols <- fti_results_ancestor$column[fti_results_ancestor$influential == TRUE]
  print(influential_cols)
}
```
## Audit to verify if the chromosomal "maps" match between ClinVar and GWAS
```{r}
library(jsonlite)
library(dplyr)
library(stringr)

# --- THE AUDIT ---
check_genome_build <- function(Variant_data, GWAS_data, genai_obj) {
  # Sample positions for the AI to inspect
  var_pos <- na.omit(Variant_data$PositionVCF)[1:5]
  gwas_pos <- na.omit(GWAS_data$CHR_POS)[1:5]
  
  prompt <- paste0(
    "You are a bioinformatics auditor. Identify the Genome Build for these two datasets.\n",
    "Dataset A (ClinVar) sample positions: ", paste(var_pos, collapse=", "), "\n",
    "Dataset B (GWAS) sample positions: ", paste(gwas_pos, collapse=", "), "\n",
    "Return only raw JSON: {\"variant_build\": \"...\", \"gwas_build\": \"...\", \"match\": true/false}"
  )
  
  result <- txt(genai.object = genai_obj, prompt = prompt)
  return(fromJSON(gsub("```json|```", "", result)))
}
```
## Pre-join Data Cleaning

```{r}
# --- PRE-JOIN DATA CLEANING ---

# 1. Standardize rsIDs (Extract only the numbers)
# Assumes 'SNPS' in Accession and 'RS# (dbSNP)' in Variant
gwas_accession$rs_num <- str_extract(gwas_accession$SNPS, "[0-9]+")
Variant_data$rs_num <- str_extract(Variant_data$`RS# (dbSNP)`, "[0-9]+")

# 2. Standardize Chromosomes (Ensure they are characters)
gwas_accession$chr_clean <- as.character(gwas_accession$CHR_ID)
Variant_data$chr_clean <- as.character(Variant_data$Chromosome)

# 3. Standardize Positions (Ensure they are numeric)
gwas_accession$pos_clean <- as.numeric(gwas_accession$CHR_POS)
Variant_data$pos_clean <- as.numeric(Variant_data$PositionVCF)

# --- THE AUDIT EXECUTION ---
build_info <- check_genome_build(Variant_data, gwas_accession, google_obj)

# --- THE FINAL JOIN ---
if(build_info$match == FALSE) {
  message("Warning: Genome builds do not match (", build_info$variant_build, " vs ", build_info$gwas_build, ").")
  message("Joining primarily by RSID to maintain data integrity.")
  
  # Fallback: Join only on RSID
  final_master_df <- gwas_accession %>%
    left_join(Variant_data, by = "rs_num")
    
} else {
  message("Builds match (", build_info$variant_build, ").")
  message("Proceeding with high-precision Dual-Key Join (RSID + Position).")
  
  # Precision Join: Match on RSID AND Chromosome AND Position
  final_master_df <- gwas_accession %>%
    left_join(Variant_data, by = c("rs_num", "chr_clean", "pos_clean"))
}

# --- FINAL POPULATION FILTER ---
# Filter for Black/African ancestry studies from your Ancestor dataset
final_filtered_df <- final_master_df %>%
  inner_join(gwas_ancestor, by = "STUDY") %>%
  filter(grepl("African|Black", `BROAD ANCESTRAL CATEGORY`, ignore.case = TRUE))

cat("Workflow Complete. Final rows for analysis:", nrow(final_filtered_df), "\n")
```
*Perform a merge on all three datasets 
```{r}
# 1. Get the new logic
logic <- get_dual_join_logic(fti_results_variant, fti_results_accession, fti_results_ancestor, google_obj)

# 2. Study-level Join (Ancestor to Accession)
study_merged <- inner_join(gwas_accession, gwas_ancestor, by = logic$anc_to_acc)

# 3. Data Cleaning for Dual-Key Match
# Ensure Chromosomes are characters (to avoid 'X' vs 23 issues)
study_merged$chr_clean <- as.character(study_merged[[logic$chr_keys[1]]])
Variant_data$chr_clean <- as.character(Variant_data[[logic$chr_keys[2]]])

# Ensure Positions are numeric
study_merged$pos_clean <- as.numeric(study_merged[[logic$pos_keys[1]]])
Variant_data$pos_clean <- as.numeric(Variant_data[[logic$pos_keys[2]]])

# Extract RSID numbers
study_merged$rs_num <- str_extract(study_merged[[logic$rs_keys[1]]], "[0-9]+")
Variant_data$rs_num <- str_extract(Variant_data[[logic$rs_keys[2]]], "[0-9]+")

# 4. PERFORM THE DUAL-KEY JOIN
# This matches on RSID OR (Chromosome + Position)
final_master_df <- study_merged %>%
  left_join(Variant_data, by = c("rs_num", "chr_clean", "pos_clean"))

cat("Dual-Key Join Complete. Checking for matches...")
```

I need to run some simple plots for the categorical and continous variables to better understand the data that I have. Then I can resume the filter and joining. 

When order matters = Ordinal 
When order doesn't matter = Nominal 



Use histogram for ordered set of continous because it creates 'bins' for the values
- Shows you distrubition and outliers and if your data is skewed (be careful different bins can tell different stories)

Questions to ask about this data:
- Which values are most common? Why
- Which values are rare? Whhy? 
- Can you see any unsuals patterns? What might explain them? 
- Do these visualizatiolns match my expectations? 
- How are the observations of each cluster similar to each other? 
- Missing values:
  1. Investigate 
  2. Meaningless - drop them 
  3. Change them to null values 
  
Measuring 'covariation' (variable to variable, observation to observation, variable to observation)
- Boxplot