##Install and load Gemini into workspace

```{r}
library(httr2)
library(jsonlite)

#install.packages("GenAI")
library(GenAI)
usethis::edit_r_environ()
api <- Sys.getenv("GOOGLE_API_KEY")
# Setup
google <- genai.google(api = api, model ="gemini-2.5-flash", version="v1")

# 3. Use the txt() function instead of gen()
response <- txt(genai.object = google, 
                prompt = "Write a haiku about R programming.")

# 4. View the result
cat(response)


```
##Load Data into workspace

```{r}
library(tidyverse)
Variant_data <- read_tsv("variant_summary.txt", col_names = TRUE)
gwas_accession <- read_tsv("gwas-catalog-download-associations-alt-full.tsv",
                 col_types = cols(.default = "c"))
gwas_ancestor <- read_tsv("gwas-catalog-v1.0.3.1-ancestries-r2026-02-16.tsv",
                 col_types = cols(.default = "c"))
colnames(gwas_accession)
```
## Perform Feature Type Inference on columns in Each dataset

*Vairant_data -> ClinVar
```{r}
library(jsonlite)

infer_variant_column_types <- function(Variant_data, genai_obj) {
  
  # 0. Drop irrelevant columns before analysis
  cols_to_drop <- c(
    # Somatic / Oncogenicity block
    "SomaticClinicalImpact",
    "SomaticClinicalImpactLastEvaluated",
    "ReviewStatusClinicalImpact",
    "Oncogenicity",
    "OncogenicityLastEvaluated",
    "ReviewStatusOncogenicity",
    # VCF formatting columns
    "PositionVCF",
    "ReferenceAlleleVCF",
    "AlternateAlleleVCF",
    # Aggregate SCV fields
    "SCVsForAggregateGermlineClassification",
    "SVCsForAgreggateSomaticClinicalImpact",
    "SCVSForAggregatedOncogenicityClassification",
    # Testing registry
    "TestedInGTR"
  )
  
  Variant_data <- Variant_data %>%
    select(-any_of(cols_to_drop))
  
  # 1. Sample the data
  col_samples <- sapply(names(Variant_data), function(col) {
    vals <- na.omit(Variant_data[[col]])[1:3]
    if(length(vals) == 0) vals <- "All NA"
    paste0(col, ": [", paste(vals, collapse = ", "), "]")
  })
  
  # 2. Build the prompt
  prompt <- paste0(
    "You are a genomic data expert analyzing a ClinVar dataset.\n",
    "Classify each column as: Numerical, Categorical, Datetime, Ignorable_ID, Formatted_ID, Sentence, URL, List, or Text.\n",
    "CRITICAL: Flag 'influential' as true if the column contains data relevant to identifying risks for:\n",
    "1. Type 1 Diabetes (e.g., HLA alleles, INS, PTPN22 genes).\n",
    "2. Food-related Anaphylaxis/Allergies (e.g., IgE response genes, FLG, or phenotype descriptions).\n",
    "Return only raw JSON: [{\"column\": \"...\", \"type\": \"...\", \"influential\": true/false}]\n\n",
    paste(col_samples, collapse = "\n")
  )
  
  # 3. Call Gemini
  result <- txt(genai.object = genai_obj, prompt = prompt)
  
  # 4. Clean and Parse
  clean_json <- gsub("```json|```", "", result)
  
  tryCatch({
    fromJSON(clean_json)
  }, error = function(e) {
    message("JSON Parsing Failed. Raw output: ", result)
    return(NULL)
  })
}

# --- EXECUTION ---

# 1. Initialize (Paste your API key between the quotes)
google_obj <- genai.google(
  api = api, 
  model = "gemini-2.5-flash", 
  version = "v1"
)

# 2. Run the inference
# Make sure your 'Variant_data' or GWAS data is loaded before running this
fti_results_variant <- infer_variant_column_types(Variant_data, google_obj)

# 3. Filter and Print
if (!is.null(fti_results_variant)) {
  influential_cols <- fti_results_variant$column[fti_results_variant$influential == TRUE]
  cat("Influential Columns Found:\n")
  print(influential_cols)
}
```
 *gwas_ancestory -> GWAS, Ancestry
 
parse the sample size strings for a more precise percentage of black participants. I am shooting for atleast 70%
 
```{r}
library(stringr)

# Now run the percentage calculation on the unfiltered data
gwas_ancestor_pct <- gwas_ancestor %>%
  mutate(group_n = as.numeric(str_replace_all(`NUMBER OF INDIVIDUALS`, ",", ""))) %>%
  group_by(`STUDY ACCESSION`) %>%
  mutate(total_n = sum(group_n, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    is_african = str_detect(
      `BROAD ANCESTRAL CATEGORY`,
      "African American|Afro-Caribbean|African unspecified|Sub-Saharan African"
    ),
    african_n = ifelse(is_african, group_n, 0)
  ) %>%
  group_by(`STUDY ACCESSION`) %>%
  mutate(
    study_pct_african = sum(african_n, na.rm = TRUE) / max(total_n, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  # NOW filter to studies where African ancestry is at least 70%
  filter(study_pct_african >= 0.70) %>%
  # And keep only the African ancestry rows from those studies
  filter(is_african == TRUE)

gwas_ancestor_pct

gwas_ancestor_pct %>%
  summarise(n_studies = n_distinct(`STUDY ACCESSION`))

#write_tsv(gwas_ancestor_pct, 
#          "C:/Data Wrangling/Data Wrangling/Final #Project/gwas_filtered_black_70pct.tsv")
```

I need to remove the columns I don't want to keep from gwas_ancestor_pct so that when I join it there will be less duplications. 

```{r}
col1 <- colnames(gwas_ancestor_pct)
col2 <- colnames(gwas_accession)
col1
col2

gwas_accession <- gwas_accession %>% select((-c(2,3,4)))

colnames(gwas_accession)
```


I now need to join my filtered ancestry data with my accession data by the  study accession

```{r}
gwas_joined <- gwas_ancestor_pct %>%
  inner_join(gwas_accession, 
            by = "STUDY ACCESSION",
            relationship = "many-to-many")
colnames(gwas_joined)
nrow(gwas_joined)
n_distinct(gwas_joined$`SNPS`)
n_distinct(gwas_joined$`STUDY ACCESSION`)

gwas_joined %>% summarise(
  Rows = n(),
    Distinct_SNPs = n_distinct(SNPS),
    Distinct_Accessions = n_distinct(`STUDY ACCESSION`)
)
#get rid out duplicate columns
remove_duplicate_cols_llm <- function(df, genai_obj) {
  
  col_list <- paste(colnames(df), collapse = ", ")
  
  prompt <- paste0(
    "You are a data wrangling expert analyzing a joined GWAS dataset.\n",
    "Here are the column names after joining an ancestry file with a GWAS associations file:\n\n",
    col_list, "\n\n",
    "Identify columns that are:\n",
    "1. Redundant duplicates of another column\n",
    "2. Intermediate calculation columns no longer needed for analysis\n",
    "3. Irrelevant to a study focused on Type 1 Diabetes, food allergy/anaphylaxis,",
    " and Black population representation in GWAS\n\n",
    "Return only raw JSON: [\"column_name_1\", \"column_name_2\"] — ",
    "a flat list of column names to DROP. No explanation, no markdown."
  )
  
  # Use your working method instead of llm_call
  result <- txt(genai.object = genai_obj, prompt = prompt)
  
  # Parse the JSON list of columns to drop
  cols_to_drop <- fromJSON(gsub("```json|```", "", result))
  
  cat("Gemini recommends dropping:\n")
  print(cols_to_drop)
  
  # Drop them
  df %>% dplyr::select(-any_of(cols_to_drop))
}

# Run it
gwas_joined <- remove_duplicate_cols_llm(gwas_joined, google_obj)
colnames(gwas_joined)

```
> Comments: I am noticing disproportianate amounts between the rows snps and accessions. Claude suggest this could happen for 2 reasons:
1. The same SNP could be studied in multiple different studies
2. The same SNP could be associated with multiple traits

I could explore this by: 

1. filter out the data only to repeated SNPs 

2. filter out the data only to repeated accessions

```{r}
# 1. Filter to only repeated SNPs
repeated_snps <- gwas_joined %>%
  group_by(SNPS) %>%
  filter(n() > 1) %>%
  ungroup() %>%
  arrange(SNPS)

repeated_snps <- repeated_snps %>% select(`STUDY ACCESSION`, `DISEASE/TRAIT`, SNPS) %>% arrange(SNPS)


nrow(repeated_snps)
n_distinct(repeated_snps$SNPS)

# 2. Filter to only repeated accessions
repeated_accessions <- gwas_joined %>%
  group_by(`STUDY ACCESSION`) %>%
  filter(n() > 1) %>%
  ungroup() %>%
  arrange(`STUDY ACCESSION`)

repeated_accessions <- repeated_accessions %>% select(`STUDY ACCESSION`, `DISEASE/TRAIT`, SNPS) %>% arrange(SNPS)

nrow(repeated_accessions)
n_distinct(repeated_accessions$`STUDY ACCESSION`)


```
> Comments: 
* Repeated SNPs across different studies = replication = strengthens significance
* Repeated SNPs across different traits = shared loci = your gold data
* Repeated accessions = one study found multiple SNPs = normal GWAS output

Now that I have done that I can try to explore SNPS with T1D and allergy traits even though I don't expect to find much
```{r}
# Find your shared loci
t1d_snps <- gwas_joined %>%
  filter(str_detect(`DISEASE/TRAIT`, 
         regex("type 1 diabetes|T1D|IDDM|autoimmune diabetes", 
               ignore_case = TRUE))) %>%
  pull(SNPS)

allergy_snps <- gwas_joined %>%
  filter(str_detect(`DISEASE/TRAIT`, 
         regex("anaphylaxis|allergy|mastocytosis|asthma|eczema|rhinitis|atop", 
               ignore_case = TRUE))) %>%
  pull(SNPS)

# The intersection — your core finding
shared_snps <- intersect(t1d_snps, allergy_snps)

cat("Shared SNPs between T1D and Allergy phenotypes:\n")
print(shared_snps)
cat("\nTotal shared:", length(shared_snps))
```
 So I expected that to return 0 instead I will seel the unique disease/traits
 
```{r}
# See all unique disease traits in your Black-filtered GWAS data
gwas_joined %>%
  count(`DISEASE/TRAIT`, sort = TRUE) %>%
  print(n = 50) 
```
It appears that T1D and allergy traits are barried in the frequency list of diesease and traits 
```{r}
# Search specifically for your phenotypes
gwas_joined %>%
  filter(str_detect(`DISEASE/TRAIT`,
         regex("type 1|T1D|anaphyl|food allerg|mastocyt|atop|rhinitis|eczema|asthma",
               ignore_case = TRUE))) %>%
  count(`DISEASE/TRAIT`, sort = TRUE) %>%
  print(n = 50)
```

Let's try the shared loci again 
```{r}
# Run your shared loci search again with expanded allergy terms
t1d_snps <- gwas_joined %>%
  filter(str_detect(`DISEASE/TRAIT`,
         regex("type 1 diabetes|T1D", ignore_case = TRUE))) %>%
  pull(SNPS)

allergy_snps <- gwas_joined %>%
  filter(str_detect(`DISEASE/TRAIT`,
         regex("atopic dermatitis|asthma|anaphylactic|allergic march|eczema",
               ignore_case = TRUE))) %>%
  pull(SNPS)

shared_snps <- intersect(t1d_snps, allergy_snps)
cat("Shared SNPs:", length(shared_snps))
```

Let's check if shared SNPs exist in the unfiltered GWAS data
```{r}
# Run the same search on your UNFILTERED gwas data
t1d_snps_all <- gwas_accession %>%
  filter(str_detect(`DISEASE/TRAIT`,
         regex("type 1 diabetes|T1D", ignore_case = TRUE))) %>%
  pull(SNPS)

allergy_snps_all <- gwas_accession %>%
  filter(str_detect(`DISEASE/TRAIT`,
         regex("atopic dermatitis|asthma|anaphylactic|allergic march",
               ignore_case = TRUE))) %>%
  pull(SNPS)

shared_snps_all <- intersect(t1d_snps_all, allergy_snps_all)
cat("Shared SNPs in full dataset:", length(shared_snps_all))
```
> Comments: 
* Full unfiltered GWAS data → 30 shared SNPs between T1D and allergy phenotypes
* Black ancestry filtered data (70%+) → 0 shared SNPs

This is a REALLY big discovery:
Shared genetic loci between T1D and allergic phenotypes do exist in the literature
Those same loci have never been studied in Black populations at a meaningful scale

Next, I want to dig deeper into the 30 SNPs that were caught. 

```{r}
# Get full details on your 30 shared SNPs
shared_snps_detail <- gwas_accession %>%
  filter(SNPS %in% shared_snps_all) %>%
  select(SNPS, `DISEASE/TRAIT`, `MAPPED_GENE`, 
         `REPORTED GENE(S)`, `P-VALUE`, 
         `RISK ALLELE FREQUENCY`,
         `INITIAL SAMPLE SIZE`) %>%
  arrange(SNPS)

# View them
print(shared_snps_detail, n = 50)

# Check which genes they map to
shared_snps_detail %>%
  count(`MAPPED_GENE`, sort = TRUE)

# Check which traits they span
shared_snps_detail %>%
  count(`DISEASE/TRAIT`, sort = TRUE)
```

I want to add my clinvar variant summary data the accession dataset with the 30 SNPS as a clinical validation layer. 

```{r}
# ============================================================
# CLINVAR VALIDATION OF 30 SHARED SNPs
# ============================================================

# ---- 1. Clean ClinVar - Drop Irrelevant Columns ----
cols_to_drop <- c(
  "SomaticClinicalImpact",
  "SomaticClinicalImpactLastEvaluated",
  "ReviewStatusClinicalImpact",
  "Oncogenicity",
  "OncogenicityLastEvaluated",
  "ReviewStatusOncogenicity",
  "PositionVCF",
  "ReferenceAlleleVCF",
  "AlternateAlleleVCF",
  "SCVsForAggregateGermlineClassification",
  "SVCsForAgreggateSomaticClinicalImpact",
  "SCVSForAggregatedOncogenicityClassification",
  "TestedInGTR"
)

Variant_data_clean <- Variant_data %>%
  select(-any_of(cols_to_drop))

# ---- 2. Match ClinVar entries to your 30 shared SNPs ----
clinvar_rsids <- Variant_data_clean %>%
  mutate(SNPS = paste0("rs", `RS# (dbSNP)`)) %>%
  filter(SNPS %in% shared_snps_all)

# ---- 3. Check coverage ----
cat("Shared SNPs with ClinVar entries:", n_distinct(clinvar_rsids$SNPS))
cat("\nShared SNPs NOT in ClinVar:", 
    length(setdiff(shared_snps_all, clinvar_rsids$SNPS)))

# ---- 4. Join ClinVar validation to shared SNP details ----
shared_validated <- shared_snps_detail %>%
  left_join(
    clinvar_rsids %>%
      dplyr::select(
        SNPS, GeneSymbol, ClinicalSignificance,
        ReviewStatus, NumberSubmitters, PhenotypeList
      ) %>%
      # Keep the most reviewed ClinVar entry per SNP
      group_by(SNPS) %>%
      slice_max(as.numeric(NumberSubmitters), n = 1, with_ties = FALSE) %>%
      ungroup(),
    by = "SNPS",
    relationship = "many-to-many"
  )

# Check results
cat("Rows in validated table:", nrow(shared_validated))
cat("\nDistinct SNPs:", n_distinct(shared_validated$SNPS))

# ---- 5. View final validated table ----
shared_validated %>%
  select(
    SNPS, `DISEASE/TRAIT`, `MAPPED_GENE`,
    ClinicalSignificance, ReviewStatus,
    NumberSubmitters, PhenotypeList
  ) %>%
  print(n = 50)


```
> Comments: 
* 2 SNPs have ClinVar entries
* 28 SNPs are NOT in ClinVar
* 30 distinct SNPs total 

Overall findings:
2 SNPs have ClinVar entries
28 SNPs are NOT in ClinVar
30 distinct SNPs total 

The 2 validated SNPs 

```{r}
# Isolate the 2 SNPs that DO have ClinVar entries
clinvar_confirmed <- shared_validated %>%
  filter(!is.na(ClinicalSignificance)) %>%
  dplyr::select(
    SNPS, `DISEASE/TRAIT`, `MAPPED_GENE`,
    GeneSymbol, ClinicalSignificance,
    ReviewStatus, NumberSubmitters, PhenotypeList
  )

print(clinvar_confirmed)
```
> Comments: 
The two Genes:
IL7R (Interleukin-7 Receptor)
This is a well established T1D susceptibility gene — IL7R variants affect T cell development and survival, which is directly relevant to the autoimmune destruction of pancreatic beta cells in T1D. But crucially IL7R is also implicated in allergic inflammation and atopic conditions because it plays a role in regulating the balance between immune tolerance and hypersensitivity. A shared SNP here makes strong biological sense.
RPS26 (Ribosomal Protein S26)
This one is particularly interesting — RPS26 sits in the chromosome 12q13 region and has been associated with T1D susceptibility through effects on INS (insulin gene) expression. More recently it has been implicated in immune cell function broadly. Its appearance as a shared locus between T1D and allergy phenotypes is a novel finding worth highlighting.

Full clinical details here:
```{r}
clinvar_confirmed %>%
  select(
    SNPS, `DISEASE/TRAIT`, GeneSymbol,
    ClinicalSignificance, ReviewStatus,
    NumberSubmitters, PhenotypeList
  ) %>%
  print(n = 50)
```
Deep dive into the specific allergy and T1D traits 

```{r}
shared_snps_detail %>%
  filter(SNPS %in% clinvar_confirmed$SNPS) %>%
  select(SNPS, `DISEASE/TRAIT`, `MAPPED_GENE`) %>%
  arrange(SNPS)

```
> Comments: Now I can see what the SNP is associated with: 
rs6897932 → IL7R
This SNP is associated with:
Type 1 diabetes ✅
Asthma ✅
Multiple sclerosis
Primary biliary cholangitis
Systemic lupus erythematosus
IL7R protein levels
rs705704 → RPS26
This SNP is associated with:
Type 1 diabetes ✅
Asthma ✅
Body mass index
Metabolic syndrome
Forced expiratory volume
Medication use (thyroid preparations)



Summary Narrative 
```{r}
# Summarize your complete findings
cat("==============================================")
cat("\nKEY FINDINGS SUMMARY")
cat("\n==============================================")
cat("\n1. Shared T1D-Allergy SNPs in full GWAS:", length(shared_snps_all))
cat("\n2. Shared SNPs studied in Black populations:", length(shared_snps_black))
cat("\n3. Shared SNPs clinically validated in ClinVar:", n_distinct(clinvar_confirmed$SNPS))
cat("\n4. Shared SNPs never clinically catalogued:", 
    length(setdiff(shared_snps_all, clinvar_rsids$SNPS)))
cat("\n----------------------------------------------")
cat("\nClinically validated shared loci:")
cat("\n  - rs6897932 (IL7R): T1D + Asthma + Autoimmune")
cat("\n  - rs705704  (RPS26): T1D + Asthma + Metabolic")
cat("\n==============================================\n")
```
I need to run some simple plots for the categorical and continous variables to better understand the data that I have. Then I can resume the filter and joining. 

When order matters = Ordinal 
When order doesn't matter = Nominal 



Use histogram for ordered set of continous because it creates 'bins' for the values
- Shows you distrubition and outliers and if your data is skewed (be careful different bins can tell different stories)

Questions to ask about this data:
- Which values are most common? Why
- Which values are rare? Whhy? 
- Can you see any unsuals patterns? What might explain them? 
- Do these visualizatiolns match my expectations? 
- How are the observations of each cluster similar to each other? 
- Missing values:
  1. Investigate 
  2. Meaningless - drop them 
  3. Change them to null values 
  
Measuring 'covariation' (variable to variable, observation to observation, variable to observation)
- Boxplot