---
title: "Written Proposal Report"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
##Outline:##
***-Introduction and motivation***
  -Personal connection to loved one with type 1 diabetes and I want to use this project as a way to insert my way into the field 
  -I have found some fluent themes like underepsentation in the black community while yet type 1 diabetes is increasing. Black people with type 1 are more likely to go into ketoacidosis and the reason for this is inclusive with a few papers citing that it could be due to the person with type 1 and their caregivers not knowing how use the pump. I believe that with the combination of my experience perspective and studying data science I could bring something meaningful to both communities and I am wanting to pursue data science for social good so I think this is a great application to do so. 
I have mapped a lot of research being done in relation with type 1 diabetes and celliac diesease but I have  only found a few papers discussing my topic specifically. I was listening to a podcast about early onset type 1 diabetes in that podcast there was mention of a big event happening that takes type 1 from being domant  in the body to slowly but progressively overtaking the pancreas. When my brother was like two he had a severe allergic reaction like anaphylaxis level and  three years later we were able to get him a diagnosis and even then the doctors continued to say it was nothing and not test him. I know that there is a lot in that series of actions to be research I just need to find out what to start with. 
***-Related work (drawing from the required readings)***
  - From some papers I read tonight there is a handful of research being done with comparing serve allergies to type 1 and there have been foundings of a correlation the are focusing on a SPECIFIC allergen. SO I am thinking I might need to narrow my focus the tree allergies for instances BUT I also have yet to come across any looking at a development of increasing food allergies AFTER the onset of type 1 diabetes. Originally I read in a paper that the presence of type 1 and allergies were "immunologically antagonistic" because the TH1 (diabetes related) and TH2 (allergy related) are polarized in the way they interact with the immune system; essentially pulling the immune system in two different directions. However, upon further reading this appears to be oversimplification and perhaps outdated. As there is frequent studies on the correlation to type 1 and asthma and excsema and ceiliac disease. 
Papers:
[https://www.sciencedirect.com/science/article/pii/S2772829325001201] - Based on sweden participants from 1987-2021 and found associations between Asthma, allergic rhinitis, and eczema to type 1 diabetes. And states that there is evidence of TH1 and TH2 immunity being mutually "inhibitory systems" was an oversimplification. Communally there is a confidence between asthma and type 1 having a mutually relationship but there are conflicting results between allergic rhinitis or eczema and type 1. 

***-Dataset description and source***
I use ClinVar and the NCBI ecosystem as my primary data foundation 
act as the industry-standard 'bridge' between raw genetic data and clinical reality. ClinVar gives me a centralized, vetted way to map genetic variants to specific phenotypes, which is essential for ensuring my model’s inputs are clinically grounded. NCBI’s broader infrastructure (like MedGen) allows me to standardize those phenotypes using universal ontologies.Essentially, it’s the most reliable way to ensure my research is built on curated, high-confidence data rather than just fragmented literature

I use the GWAS Catalog because it is the industry standard for mapping genotype-phenotype associations at a massive scale. It provides high-quality, standardized data that makes meta-analyses and large-scale data science projects much more reliable. It is crucial for demographic research, as it allows me to integrate ancestral diversity into my models, which is essential for reducing health inequities and improving the accuracy of risk predictions across different populations.

***-Planned LLM implementation approach***
  - I am interesting in using the LLM to help me find a middle ground for the demographic filtering of the GWAS ancerstry data. Also I am going to task the LLM with joining the GWAS data with the variant data, I am nervous to do this because I feel it will take some effort in validating it. I am also curios to see if like the CodeGenWrangler LLM if I can use external knowledge to help me quantify the amount of black participants in the studies AND/OR use it to find better phenotypes and genosymbols for labeling type 1 diabetes and anaphylaxis.
  
  - Use LLMs to join GWAS subset with ClinVar subset via the RSID and the chromosone + base pair position. There may be some formatting differences so I think an LLM will be perfect for the task 
  - Then us it to find the 'influential columns' 
***-Preliminary data quality assessment***
The variant_summary is high quality as it is complete and standardized. 
***-Summary of planned work***
Current focus: READ MORE PAPERS 
Possibly adjust and restate an iterative version of my question.
Use LLM to re-employ the EDA checklist with a revised list of key words for type 1 and anaphylaxis. Use LLM to better filter GWAS ancestry data  and to join GWAS data with variant data. 
The next step beyond this depends on the amount of results that are returned to me. 

I need to validate this data with 1 external source 
I need to make a plot to try and find points that may be missing in the table 
- create expectations
- check for deviationss
The easy solution would be there simply is not correlation but in the papers I have read there most definitely is. 
- follow ups:
-do I have the right data
-do I need more data
- is the descripency between the is data being primarily focused on genomes and phentypes
- also could the issue be that my definition on anaphylaxis is wrong, can I filter for TH1 and TH2 with this data?

##Research Question: ##
What shared variants/loci exist between severe early-onset **food** allergies phenotypes and T1D, and are these loci underrepresented in Black populations in ClinVar?

My Expectations:
I expect to find a correlated relationship between early-onset severe allergic responses and the subsequent development of Type 1 Diabetes. From the Genome Wide Association Studies there was no correlations



Load and Read the `variant_summary.txt` into R environment
```{r}
library(tidyverse)
getwd()
setwd("C:/Data Wrangling/Data Wrangling/Final Project")

```
Read Data in tsv
```{r}
Variant_data <- read_tsv("variant_summary.txt", col_names = TRUE)
glimpse(Variant_data)
```
Check the packaging
1. Check for null values
```{r} 
str(Variant_data)































```
```{r}
colSums(is.na(Variant_data))
```
```{r}
sum(is.na(Variant_data))
```
2. Check number of rows 
```{r}
nrow(Variant_data)

```
3. Check number of columns 
```{r}
ncol(Variant_data)
```

Print columns
```{r}
colnames(Variant_data)
```
With the help of Gemini I have identified that my columns a combination of 4 different categories:
1. The Core identifiers and basic info
  - #AlleleID: primary key for a specific allele(specific variant) in CLinVar
  - Type: the type of variant (single nucleotide variant = SNV), deletion, insertion, duplication, etc. 
  - Name: full (human readable name) - often written in HGVS notation
  - GeneID: tells us where the gene is located (NCBI Entrez Gene ID)
  - GeneSymbol: The short official symbol for the gene 
  - HGNC_ID: ID assigned by the HUGO Gene Nomenclature Committee (office authority for human gene naming)
2. Clinical Significance 
  - ClinicalSignificance: full 'clinical' interpretation of the variant (Pathogenic, Benign, Uncertain significance). *if submitters disagree multiple terms with be included
  - ClinSigSimple: Numeric version of ClincalSignificance 
  - LastEvaluated: most recent date that submitter review/evaluated the ClinicalSignificance of that specific variance
  - TeviewStatus: reflects how well supported the classification is ('no assertion criteria provided' to 'practice guideline'. More stars = More confidence)
  - NumberSubmitters: reflects how many independent labs or organizations have submitted evidence about this variant to ClinVar
  - Guidelines: list any 'official' clinicalguidelines that reference this variant
3. Disease/Phenotype
  - PhenotypeIDS: Database IDS (from MedGen, OMIM, Orphanet) for the diseases(s) or conditions(s) associated with the variant
  - PhenotypeList: The names of the conditions
  - Origin: where the variant came from (germiline=inherited, somatic=acquired in a tumor, de novo=new mutation)
  - OriginSimple: simplified version of the origin
4. Genomic Location
  - Assembly: the human genome reference assembly 
  - ChromosomeAccession: the official NCBIS acession number match to the specific chromosome sequence used
  - Chromosome: the chromosome where the variant is located 
  - Start: the starting genomic coordinate of the variant on the chromosome
  - Stop: the ending genomic coordinate (same as the start for a single nucleotide variants)
  - Cytogenetic: the cytogenetic band location (used in karyotyping)
  - ReferenceAllele: the nucleotide(s) found in the reference genome at the position
  - AlternateAllele: the nucleotide(s) that differ from the references (the actual variant sequence)
5. Cross-Database References
  - RS#(dbSNP): the number from the NCBI's database of known variants
  - nsv/esv(dbVAR): structural variant ID from dbVar, NCBI's database for larger genomic variants (i.e. copy number variants)
  - RCVaccession: tied to a specific variant-condition pair (one variant can have many RVC's for different diseases) 
  - VariationID: this ID is independent of the condition. Used ot look it up directly on the CinVar website
  - OtherIDS: Additional cross-references to other databases
6. VCF Fields
  - PositionVCF:chromosomal position as it would appear in a standard VCF (Variant Call Format) file
  - ReferenceAlleleVCF: reference allele formatted specifically for VCF standards 
  - AlternateAlleleVCF: alternate allele formatted for VCF standards
7. Submission & Testing
  - TestedInGTR: Y or N for whether this variant is included in a genetic test registered in NCBI's Genetic Testing Registry (GTR)
  - SubmitterCategories: the types of organizations that submitted data (clinical laboratory, research, expert panel)
8. Somatic/Oncogenicity Fields
  - SomaticClinicalImpact: the clinical impact of the variant when found in a tumor (somatically) - like "Tier 1 - Strong Clinical Significance"
  - SomaticClinicalImpactLastEvaluated: the date the somatic clinical impact was last reviewed
  - ReviewStatusClinicalImpact: the confidence level specifically for the somatic clinical impact classification
  - Oncogenicity: is the variant consider 'cancer-causing' (oncogenic) when its a tumor (somatic)
  - OncogenicityLastEvaluated: the date the oncogenicity classification was last reviewed
  - ReviewStatusOncogenicity: confidence level for the oncogenicity classification
9. AggregateSCVFields
  - SCVsForAggregateGermlineClassification: the list of individual SCVs that contributed to the overall germline pathogenicity classification
  - SVCsForAgreggateSomaticClinicalImpact: the list of SCVs contributing to the somatic clinical impact classification
  - SCVSForAggregatedOncogenicityClassification: the list of SCVs contributing to the oncogenicity classification




Print the head
```{r}
head(Variant_data)
```

Print the tail
```{r}
table(sapply(Variant_data, class))
```
Filter down to type 1 data using Keywords: type 1 diabetes, T1D, IDDM, insulin-dependent diabetes

Widen the search and then also filtered to find genesymbols associated with type 1 diabetes to further sift out any noise within the data
```{r}
t1d_variants <- Variant_data %>%
  filter(str_detect(PhenotypeList, regex("type 1 diabetes|T1D|IDDM|insulin-dependent diabetes|diabetes mellitus|type 1 diabetes mellitus 20|monogenic autoimmune diabetes|autoimmune diabetes|neonatal diabetes mellitus|immune related diabetes|MODY|maturity-onset diabetes of the young|maturity onset diabetes mellitus in young", ignore_case = TRUE))) %>% filter(str_detect(GeneSymbol, regex("HLA-DRB1|HLA-DQA1|HLA-DQB1|INS|CTLA4|PTPN22|HNF1A|GCK|HNF4A", ignore_case = TRUE)))

# Check how many variants 
nrow(t1d_variants)

# Preview
glimpse(t1d_variants)
```

Check the Phenotypes and their counts <----- Human Version Results
```{r}
t1d_variants %>%
  count(PhenotypeList, sort = TRUE)
```

Severe Anaphalaxus filter
```{r}
anaphylaxis <- Variant_data %>%
  filter(str_detect(PhenotypeList, regex("mastocytosis|Hereditary alpha-tryptasemia|HαT|Idiopathic anaphylaxis|Mast cell activation syndrome", ignore_case = TRUE))) %>% filter(str_detect(GeneSymbol, regex("KIT|TPSAB1|FCER1A|(IL4&IL3)", ignore_case = TRUE)))

# Check how many variants 
nrow(anaphylaxis)

# Preview
glimpse(anaphylaxis)

t1d_variants %>% inner_join(anaphylaxis, by= 'GeneID')

```
```{r}
anaphylaxis %>%
  count(PhenotypeList, sort = TRUE)
```
Official Trait List:
- Phenotypes:
    Anaphylaxis: 
      mastocytosis
      Hereditary alpha-tryptasemia
      HαT
      Idiopathic anaphylaxis|Mast cell activation syndrome
    T1D: 
      type 1 diabetes
      T1D
      IDDM
      insulin-dependent diabetes
      diabetes mellitus
      type 1 diabetes mellitus 20
      monogenic autoimmune diabetes
      autoimmune diabetes|neonatal diabetes mellitus
      immune related diabetes|MODY|maturity-onset diabetes of the young|maturity onset diabetes mellitus in young
- GeneSymbols:
    Anaphylaxis: 
      KIT
      TPSAB1
      FCER1A
      IL4&IL3
    T1D: 
      HLA-DRB1
      HLA-DQA1
      HLA-DQB1
      INS
      CTLA4
      PTPN22
      HNF1A
      GCK
      HNF4A

Repeat filter process by via LLM query 
```{r}

```


Download GWAS acession dataset for ...
```{r}
getwd()
gwas <- read_tsv("C:/Data Wrangling/Data Wrangling/Final Project/gwas-catalog-download-associations-alt-full.tsv",
                 col_types = cols(.default = "c"))
glimpse(gwas)
table(sapply(gwas, class))
sum(is.na(gwas))
```

Preform tidy data task
```{r}
head(gwas)
```
Download ancestor GWAS dataset to get the demogrpahic information of the studies. 

```{r}
gwas_ancestor <- read_tsv("C:/Data Wrangling/Data Wrangling/Final Project/gwas-catalog-v1.0.3.1-ancestries-r2026-02-16.tsv",
                 col_types = cols(.default = "c"))
glimpse(gwas_ancestor)
gwas_ancestor %>%
  group_by(`BROAD ANCESTRAL CATEGORY`)%>%
  count()

table(sapply(gwas, class))

sum(is.na(gwas))
```
Filter for GWAS studies that atleast inclue "African American or Afro-Caribbean|African unspecified" in the Broad Ancestral Category 

```{r}
library(stringr)

ancestry_processed <- gwas_ancestor %>%
  mutate(has_african_ancestry = str_detect(
    `BROAD ANCESTRAL CATEGORY`, 
    "African American or Afro-Caribbean|African unspecified"
  ))

# Now you can filter your main dataset for these studies
african_ancestry_studies <- ancestry_processed %>%
  filter(has_african_ancestry == TRUE)

glimpse(african_ancestry_studies)
```
Filter for observations that ONLY include black participants.

```{r}
gwas_ancestor <- gwas_ancestor%>% filter(`BROAD ANCESTRAL CATEGORY`== 'African American or Afro-Caribbean' | `BROAD ANCESTRAL CATEGORY`== 'African American' | `BROAD ANCESTRAL CATEGORY`== 'African unspecified'|`BROAD ANCESTRAL CATEGORY`== 'Afro-Caribbean' |`BROAD ANCESTRAL CATEGORY`== 'Sub-Saharan African'| `BROAD ANCESTRAL CATEGORY`== 'African unspecified, African American or Afro-Caribbean'| `BROAD ANCESTRAL CATEGORY`=='African American or Afro-Caribbean, NR'| `BROAD ANCESTRAL CATEGORY`== 'African American or Afro-Caribbean, African unspecified'|`BROAD ANCESTRAL CATEGORY`=='	African unspecified, NR'|`BROAD ANCESTRAL CATEGORY`=='NR, African unspecified'|`BROAD ANCESTRAL CATEGORY`== '	Sub-Saharan African, African American or Afro-Caribbean'|`BROAD ANCESTRAL CATEGORY`== 'Sub-Saharan African, African unspecified')

glimpse(gwas_ancestor)
```
Join both GWAS datasets 
```{r}
joined_gwas <- gwas_ancestor %>%
  inner_join(gwas)
joined_gwas

colnames(joined_gwas)

glimpse(joined_gwas)

table(sapply(joined_gwas, class))

sum(is.na(gwas))
```
proceed with the following steps in R:

1. Filter by Disease/Trait: Use the DISEASE/TRAIT column to extract rows relevant to Type 1 Diabetes and your specific severe allergic phenotypes (e.g., mastocytosis, HαT, anaphylaxis).
2. Filter by Gene: Use the REPORTED GENE(S) column to narrow down to the genes you identified (KIT, TPSAB1, HLA-DRB1, etc.).
3. Identify Shared Loci: Once you have your T1D-subset and your allergy-subset, look for overlapping SNPs or MAPPED_GENE values. A shared variant in the HLA region, for instance, would be a strong candidate for a shared genetic locus.
4. Analyze Ancestry: Since you already filtered your ancestry data, you can now check the INITIAL SAMPLE DESCRIPTION or ADDITIONAL ANCESTRY DESCRIPTION columns to see the diversity of the cohorts where these shared variants were identified. If you notice a high concentration of studies in European cohorts and very few in Black populations for these shared loci, you have empirical support for your hypothesis regarding underrepresentation.




```{r}
#t1d filter
t1d_subset <- joined_gwas %>% filter(str_detect(`DISEASE/TRAIT`, "type 1 diabetes"))
#Severe allergy filter
allergy_subset <- joined_gwas %>% filter(str_detect(`DISEASE/TRAIT`, "mastocytosis|anaphylaxis|allergy"))

# Find shared variants
shared_loci <- inner_join(t1d_subset, allergy_subset, by = "SNPS")

glimpse(t1d_subset)
glimpse(allergy_subset)
shared_loci
```
So here I found that in my current filter statements there is no overlap. I am going to broaden my filters 

SHOW THIS TO DR.AZIZ. ASK IF ITS NECESSARY TO INCLUDE THE VARCHAR DATA???

```{r}
# Expanded the search
#t1d filter
t1d_subset <- joined_gwas %>% filter(str_detect(`DISEASE/TRAIT`, "type 1 diabetes|T1D|IDDM|insulin-dependent diabetes|diabetes mellitus|type 1 diabetes mellitus 20|monogenic autoimmune diabetes|autoimmune diabetes|neonatal diabetes mellitus|immune related diabetes|MODY|maturity-onset diabetes of the young|maturity onset diabetes mellitus in young"))
#severe allergy filter
allergy_subset <- joined_gwas %>% filter(str_detect(`DISEASE/TRAIT`, "mastocytosis|Hereditary alpha-tryptasemia|HαT|Idiopathic anaphylaxis|Mast cell activation syndrome|mastocytosis|anaphylaxis|allergy"))

# Find shared variants
shared_loci <- inner_join(t1d_subset, allergy_subset, by = "SNPS")

glimpse(t1d_subset)
glimpse(allergy_subset)
shared_loci
```
OKay, next step join variant_data with the t1d and anaphylaxis to the joined_gwas data and see if there is overlapp there???




Do inferential work of cross-refencing finds with specialized black juvenile dataset
The workflow:
1. Select a specific disease (e.g., childhood asthma or early-onset diabetes).

2. Use a curated database (like the GWAS Catalog) to find variants already associated with that disease in Black populations.

3. Perform functional enrichment (using tools like DAVID or Reactome) on those gene lists to see what biological processes they are involved in.

4. Connect these processes back to known social or environmental stressors—this is where your unique perspective as a data scientist with a passion for advocacy comes in.

Filter variant_summary.txt to genomes associated with type diabetes


Use those genomes to  cross-reference 



Schedule:

Weeks 1-2 — Data Collection & Integration

Pull your ClinVar variant_summary, download the relevant gnomAD African ancestry population frequency data, and query the GWAS Catalog for known allergy and T1D associated loci. Define your phenotype scope clearly — decide exactly which ClinVar phenotype terms you're treating as "early-onset severe allergic response" and which as "Type 1 Diabetes." This definitional work is critical and worth doing carefully upfront.


Weeks 3-4 — LLM-Assisted Data Cleaning & Wrangling
This is where you leverage LLMs to standardize phenotype labels, harmonize inconsistent ClinVar terminology, flag low-quality entries, and merge your datasets. This is also where you identify and subset your African/African-American ancestry variants specifically from gnomAD. Documenting this process carefully is important because the LLM-assisted cleaning pipeline itself is a methodological contribution worth writing about.


Weeks 5-7 — Machine Learning Exploration & Analysis

This is your analytical core. Depending on your specific research questions this could include clustering variants by shared functional pathways between allergy and T1D, identifying ancestry-specific variant patterns, or building a classifier that flags variants likely to be underclassified in Black populations due to representation gaps. You don't need all of these — pick one or two that best answer your central question.


Weeks 8-9 — Interpretation & Writing

Connecting your findings back to the biological story, the representation gap narrative, and the methodological innovation of using LLMs in the pipeline. This is also where you frame your limitations and future directions honestly.


Week 10 — Buffer

Always build in buffer. Something will take longer than expected.
